---
title: "Nosocomial Outbreak Reconstruction"
subtitle: "Bayesian inference of hospital transmission trees from incomplete observations"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: true
---

```{r}
#| label: setup
#| include: false
source(here::here("scripts", "_load_project.R"))
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Background

Hospital-acquired (nosocomial) infections are rarely observed in their
entirety. A patient may be infected days before a diagnostic test turns
positive, and infected patients who are discharged or never tested are
invisible to standard surveillance. The result is a fragmented picture of the
outbreak: we observe a subset of cases and, for each, only the day the test was
returned positive—not the day of infection, and not who the infecting patient
was.

This document walks through three use cases of the simulation and inference
framework:

1. **Simulating** a synthetic outbreak on a two-ward hospital and exploring its
   dynamics.
2. **Reconstructing** the transmission tree from the incomplete observed data
   via Markov chain Monte Carlo (MCMC).
3. **Comparing** reconstruction accuracy with and without genomic information.

---

## The Model {#sec-model}

### Hospital structure

The hospital is modelled as a hierarchy of spatial units: beds are nested
inside rooms, rooms inside wards, and wards inside the hospital. A two-ward
layout is used throughout:

```{r}
#| label: hospital-layout
#| echo: false
cat(sprintf(
  "Wards: %d  |  Rooms: %d  |  Beds: %d\n",
  length(unique(Positions$Ward)), NumRooms, NumBeds
))
```

### Transmission dynamics

Each day, a susceptible bed $i$ is exposed to the infectious pressure from
every currently infected bed $j$. The per-contact transmission rate depends on
the spatial relationship between $i$ and $j$:

$$
\beta_{ij} =
\begin{cases}
\beta_{\text{room}} & \text{if } i \text{ and } j \text{ share a room} \\
\beta_{\text{ward}} & \text{if } i \text{ and } j \text{ share a ward (different rooms)} \\
\beta_{\text{hosp}} & \text{otherwise}
\end{cases}
$$

The probability that bed $i$ escapes infection on day $t$ is the product of
escape probabilities over all infected contacts:

$$
P(\text{escape}_i \mid t) = \prod_{j \in \mathcal{I}(t)} (1 - \beta_{ij})
$$

so the probability that $i$ becomes infected on day $t$ is
$1 - P(\text{escape}_i \mid t)$.

The default transmission rates used in all examples below are
$\beta_{\text{room}} = 0.08$, $\beta_{\text{ward}} = 0.005$, and
$\beta_{\text{hosp}} = 0.0005$.

### Basic reproduction number

With $n_s$ contacts at spatial scale $s$ and daily discharge rate $p_{\text{dis}}$,
the basic reproduction number is

$$
\mathcal{R}_0 = \frac{\displaystyle\sum_{s} \beta_s \, n_s}{p_{\text{dis}}}
$$

```{r}
#| label: r0-value
#| echo: false
R0_val <- r_0(theta)
cat(sprintf("R\u2080 = %.2f  (with p_dis = %.2f, N_beds = %d)\n",
            R0_val, theta$P$Dis[[1]], NumBeds))
```

At $\mathcal{R}_0 > 1$ the deterministic SIS model predicts an endemic
equilibrium prevalence of $1 - 1/\mathcal{R}_0$.

### Genomic observations

Each transmission event accumulates a Poisson-distributed number of mutations
with mean $\mu$. For two cases $i$ and $j$ separated by $\kappa_{ij}$
transmission steps, the observed pairwise SNP distance is

$$
D_{ij} \;\sim\; \operatorname{Poisson}(\mu \cdot \kappa_{ij})
$$

### Bayesian reconstruction

Given observed case records $\mathbf{t}_{\text{obs}}$ (positive-test times and
bed assignments) and a genetic distance matrix $\mathbf{D}$, the MCMC targets
the joint posterior over infection times $\mathbf{t}_{\text{inf}}$ and
ancestor assignments $\mathbf{a}$:

$$
\pi\!\left(\mathbf{a},\,\mathbf{t}_{\text{inf}} \;\middle|\; \mathbf{D},\,\mathbf{t}_{\text{obs}}\right)
\;\propto\;
\underbrace{\mathcal{L}(\mathbf{D} \mid \mathbf{a},\,\mathbf{t}_{\text{inf}})}_{\text{genomic}}
\;\cdot\;
\underbrace{\mathcal{L}(\mathbf{t}_{\text{obs}} \mid \mathbf{t}_{\text{inf}})}_{\text{testing}}
\;\cdot\;
\underbrace{p_{\text{contact}}(\mathbf{a} \mid \mathbf{t}_{\text{inf}})}_{\text{spatial prior}}
\;\cdot\;
\underbrace{\pi_{\text{import}}(\mathbf{a})}_{\text{import prior}}
$$

The spatial prior $p_{\text{contact}}$ assigns higher prior probability to
ancestor–descendant pairs that share the same room or ward and whose inferred
infection times are temporally plausible.

---

## Use Case 1 — Simulating an Outbreak {#sec-sim}

### Setup and simulation

We seed a single infected patient in the first bed and run the model for 45
days.

```{r}
#| label: sim-setup
theta_demo <- local({
  th        <- theta
  th$P$Init <- c(1, rep(0, NumBeds - 1), rep(0, NumRooms))
  th
})

set.seed(81369)
outbreak <- simulate_outbreak(45, theta_demo)

cat(sprintf(
  "Total infections: %d  |  Observed (tested positive): %d\n",
  nrow(outbreak$CaseRec), nrow(outbreak$ObsRec)
))
```

::: {.callout-note}
Only cases that test positive *before* discharge enter the observed data.
The remaining infected patients form the **hidden** portion of the outbreak.
:::

### Prevalence dynamics

The simulated prevalence (proportion of beds that are infected) is overlaid
against the deterministic SIS endemic equilibrium $1 - 1/\mathcal{R}_0$.

```{r}
#| label: fig-prevalence
#| fig-cap: "Simulated daily prevalence (blue) compared to the deterministic SIS endemic equilibrium (dashed red) and the simulation mean (dotted green)."
prev <- outbreak_prevalence(outbreak)
plot_prevalence(prev)
```

With $\mathcal{R}_0 \approx `r round(r_0(theta), 2)`$ the endemic equilibrium
is around
`r round(100 * max(0, 1 - 1/r_0(theta)), 0)`% prevalence,
and the simulation tracks this closely after an initial build-up phase.

### Ground-truth transmission tree

The interactive timeline below shows every infected patient (including
unobserved cases) as a horizontal stay bar. Directed edges follow the true
transmission tree. Node colour encodes the spatial scale of each transmission
event—room, ward, or hospital-wide.

```{r}
#| label: fig-true-tree
#| fig-cap: "Ground-truth transmission tree (all cases, including unobserved). Darker edges indicate transmission within the same room."
plot_timeline(outbreak, observed = FALSE, show_all_stays = TRUE)
```

When we restrict to the **observed** cases only, the picture looks sparser:

```{r}
#| label: fig-obs-tree
#| fig-cap: "Ground-truth tree restricted to observed (test-positive) cases."
plot_timeline(outbreak, observed = TRUE, show_all_stays = TRUE)
```

---

## Use Case 2 — Bayesian Reconstruction via MCMC {#sec-mcmc}

Given only the observed case records and pairwise SNP distances, we run MCMC
to reconstruct the posterior distribution over transmission trees.

```{r}
#| label: mcmc-run
set.seed(2413)
fit <- mcmc(outbreak, N_samples = 3000, burn_in = 500)

score <- ancestry_score(
  outbreak$ObsRec$Anc2,
  fit$anc,
  outbreak$ObsRec$Adm,
  outbreak$ObsRec$PTest
)
```

### Reconstruction accuracy

```{r}
#| label: tbl-score
#| echo: false
#| tbl-cap: "Ancestry reconstruction accuracy metrics."
data.frame(
  Metric      = c("Mode accuracy", "Mean P(true ancestor)", "Mean lift", "Log skill"),
  Value       = round(c(score$mode_accuracy, score$mean_p_true,
                         mean(score$lift), score$log_skill), 3),
  Description = c(
    "Fraction of cases where the posterior-mode ancestor equals the true ancestor",
    "Average posterior probability assigned to the true ancestor",
    "Mean performance relative to a random (uniform) baseline",
    "Mean log-probability gain over the uniform baseline"
  )
) |> knitr::kable(align = c("l", "r", "l"))
```

### Posterior-mode reconstruction

Green edges are correctly inferred transmissions; red edges are errors. Edge
width reflects posterior support, and node borders indicate classification
correctness.

```{r}
#| label: fig-reconstruction
#| fig-cap: "Posterior-mode reconstruction overlaid on the ground-truth tree. Green = correct, red = incorrect."
plot_timeline_mode(outbreak, fit, color_edges = TRUE, color_nodes = FALSE)
```

### Posterior uncertainty

We can also show all edges that achieve at least 20% posterior probability,
giving a sense of uncertainty around each inferred link.

```{r}
#| label: fig-posterior
#| fig-cap: "Credible transmission edges (posterior probability ≥ 0.20)."
plot_timeline_posterior(outbreak, fit, min_prob = 0.20)
```

### MCMC convergence diagnostics

```{r}
#| label: fig-diagnostics
#| fig-cap: "MCMC convergence panel: community-import trace, autocorrelation, mean infection-time trace, ancestry raster, and per-case entropy."
#| fig-height: 9
plot_mcmc_diagnostics(fit, outbreak)
```

---

## Use Case 3 — Genomics vs. No Genomics {#sec-genetics}

Does adding the SNP distance matrix meaningfully improve reconstruction?  We
re-run MCMC on the same outbreak with `use_genetics = FALSE` and compare.

```{r}
#| label: mcmc-nogen
set.seed(8831)
fit_nogen <- mcmc(outbreak, N_samples = 3000, burn_in = 500,
                  use_genetics = FALSE)

score_nogen <- ancestry_score(
  outbreak$ObsRec$Anc2,
  fit_nogen$anc,
  outbreak$ObsRec$Adm,
  outbreak$ObsRec$PTest
)
```

```{r}
#| label: tbl-compare
#| echo: false
#| tbl-cap: "Reconstruction accuracy with and without genomic data."
data.frame(
  Metric = c("Mode accuracy", "Mean P(true ancestor)", "Mean lift", "Log skill"),
  With    = round(c(score$mode_accuracy,       score$mean_p_true,
                    mean(score$lift),           score$log_skill), 3),
  Without = round(c(score_nogen$mode_accuracy, score_nogen$mean_p_true,
                    mean(score_nogen$lift),     score_nogen$log_skill), 3)
) |> knitr::kable(align = c("l", "r", "r"),
                  col.names = c("Metric", "With genetics", "Without genetics"))
```

```{r}
#| label: fig-nogen-reconstruction
#| fig-cap: "Posterior-mode reconstruction *without* genomic data."
plot_timeline_mode(outbreak, fit_nogen, color_edges = TRUE, color_nodes = FALSE)
```

Genomic distances tend to help most when multiple plausible ancestor–descendant
pairs exist in the same room at the same time: the SNP count can then
disambiguate who infected whom.

---

## Summary

| Feature | Value |
|---|---|
| Hospital layout | 2 wards, `r NumRooms` rooms, `r NumBeds` beds |
| Simulation length | 45 days |
| $\mathcal{R}_0$ | `r round(r_0(theta), 2)` |
| Total cases | `r nrow(outbreak$CaseRec)` |
| Observed cases | `r nrow(outbreak$ObsRec)` |
| MCMC samples (post burn-in) | 3 000 |
| Mode accuracy (with genomics) | `r round(score$mode_accuracy, 3)` |
| Mode accuracy (without genomics) | `r round(score_nogen$mode_accuracy, 3)` |

The framework provides a principled way to reason about transmission chains
under the double uncertainty of partial observation and stochastic infection
dynamics. Genomic data consistently tightens the posterior, particularly in
settings with multiple infectious individuals in close spatial proximity.
